Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-10 rebuttal.tex

\documentclass[letterpaper, 12pt]{article}
\usepackage{url}
\usepackage{ctable}
\usepackage{amsmath,amssymb}      % for \AA etc
\usepackage[onehalfspacing]{setspace}
\usepackage{graphicx}
\usepackage{times}

\DeclareGraphicsExtensions{.eps,.png}
\onehalfspacing

\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}
\usepackage{anysize}

\begin{document}

I would like to thank the reviewers for their thorough comments on the original submissions. I have
prepared a revised manuscript that address the comments and my responses are detailed below.

\fbox{\textbf{Reviewer: 1}}

Recommendation: Publish after major revisions noted.

Comments:
The prediction of activity cliffs is a nice idea that would further extend activity landscape modeling. This is attempted by predicting SALI values for compound pairs on the basis of training data. Unfortunately, there are a number of technical issues to note and, more importantly, methodological concerns that would need to be carefully addressed before publication of this study might be considered.

First, a few general comments: 
(1) Please, note that there are numerous typographical errors throughout the manuscript (which is quite disturbing). Unfortunately, the presentation of the work is sloppy at different levels.       
(2) Chosen methods and descriptors should at least be referenced. For example, no references are provided for BCI and CDK.

Second, there are shortcomings at the technical level:
(3) The derivation of RF models is essentially not described (except of making reference to default parameters settings in R). It remains largely unclear how these models were developed. 
(4) The selection of training and test sets should be clearly specified in the Methods section. For example, how were the RF models trained and evaluated for which predictions of pair-wise SALI values are reported in Figure 2? It is hoped that the reported SALI values were not predicted for the training data.
(5) No support is provided for the assumption that the RF models “are relatively robust even for datasets of small size”. In fact, the author later on states several times that limited model quality might be due to the use of small compound sets. Appropriate validation will be essential here.
(6) There is only limited statistical assessment/validation of the models. Statements made in the text that R2 values are “reasonable” and the models perform “relatively well” are difficult to reconcile. Rather, Table 2 indicates that the R2 values were rather poor in several instances.
(7) It appears that there are systematic prediction errors for pairs. At low value ranges (that are not relevant for activity cliff assessment), SALI values are consistently over-predicted. By contrast, at high value ranges (that are relevant), SALI values are consistently under-predicted, in part significantly. This is not investigated, although the apparent errors at high value ranges are the perhaps most critical features of the reported RF models.    
(8) In the section “Extending a landscape”, predictions are finally reported for only three (!) molecules taken out of each data set. Carefully put, the predictions are heterogeneous. On the basis of the results reported for this little bit of evaluation, one would be hard pressed to make a case for the ability of the approach to predict activity cliffs with any certainty.

Third, there are major scientific concerns:
(9) For the prediction of activity cliffs, SALI values must be calculated for compound pairs. This is the central idea of the approach. Accordingly, fingerprint descriptors and potency values must be combined in some ways. 
(9.1.) It is nowhere stated how activity values are treated for compounds forming a pair. Are activity differences calculated at the pair level? If so, how are activity differences compared for the calculation of pair-based SALI scores - as a difference of differences? It should be noted predicted potency values in Figures 4-6 are essentially all over the place.
(9.2.) To obtain a fingerprint descriptor for a pair, the author aggregates fingerprints of individual molecules, finally, by simple averaging. For comparisons of pairs, Tanimoto similarity is then calculated for averaged fingerprints, which presents an artificial assessment of pair similarity (one can easily come up with a few hypothetical compound similarity relationships that make this assessment questionable at best). As a consequence, activity cliffs are considered for which Tanimoto similarity calculations yield values between 0.2 and 0.3! Thus, on the basis of these similarity calculations, one could never decide which level of similarity might be cliff relevant. In this respect, the descriptions of activity cliff by the author are quite telling – he speaks of “predicted” versus “true” activity cliffs, of “relatively accurate” activity cliffs, and of “most significant activity cliffs that are in fact not very significant activity cliffs in an absolute sense”, and so on. 
Furthermore, there might be many instances of activity and fingerprint similarity relationships between pairs of compounds that might yield high pair-based SALI scores dominated by either the activity or similarity component, without forming a “true” activity cliff. The conclusion of this reviewer is that it is currently not possible to capture activity cliffs with any degree of certainty on the basis of the pair calculations, as reported.
(10) It also remains largely obscure how an activity landscape should be extended following this approach by compounds for which no activity information is available. Should the activity first be predicted by an RF model and then used for pair-based SALI score calculations? One would hope not.

In summary, there will be quite a bit of work required to get this manuscript in shape for publication, both in terms of its presentation and technical content. However, as long as points (9) and (10) are not thoroughly addressed scientifically, this work should not be published.

To give the author the benefit of the doubt -and credit for the underlying idea- this reviewer calls for “Major Revisions”, although the study might as well be considered premature at this point.

Additional Questions:
Please rate the quality of the science reported in this paper (10 - High Quality/ 1 - Low Quality): 4

Please rate the overall importance of this paper to the field of chemical information or modeling (10 - High Importance / 1 - Low Importance): 6

\fbox{\textbf{Reviewer: 2}}

\textbf{1.  The manuscript is poorly written.  There are many typos, grammatical errors, missing
  references, and incomplete descriptions of methods and results.}

The text has been thoroughly proof read to remove typos and incorrect grammar. References have been
added and methods have been described more clearly.

\textbf{2.  Qualitative assessments of results are not acceptable.  The authors should give
  statistical evaluations and corresponding metrics for models and their predictions.}

\textbf{3.  The tests sets are too small to make any meaningful conclusions.}

\textbf{4.  SALI values needs to be determined for pairs of molecules - is that not the essence of
  the methodology?}

\textbf{5.  There is no discussion/direction of how to use the methodology in new applications.}
\end{document}

### 2012-07-10 paper.tex
finished proof reading the paper

\maketitle
\begin{abstract}
  The notion of activity cliffs is an intuitive approach to characterizing structural features that
  play a key role in modulating biological activity of a molecule. A variety of methods have been
  described to quantitatively characterize activity cliffs, such as SALI and SARI. However, these
  methods are primarily retrospective in nature; highlighting cliffs that are already present in the
  dataset. The current study focuses on employing a pairwise characterization of a dataset to train
  a model to predict whether a new molecule will exhibit an activity cliff with one or more members
  of the dataset. The approach is based on predicting a value for pairs of objects rather than the
  individual objects themselves (and thus allows for robust models even for small SAR datasets). We
  extracted structure-activity data for several ChEMBL assays and developed random forest models to
  predict SALI values, from pairwise combinations of molecular descriptors. The models exhibited
  reasonable RMSE's though, surprisingly, performance on the more significant cliffs tended to be
  better than on the lesser ones. Our results indicate that the predictive models are able to
  prioritize molecules in terms of their ability to activity cliffs and can be used as a way to
  extend an observed activity landscape to identify molecules that could lead to significant
  improvements in activity. Conversely, the method can be used to deprioritize molecules that are
  predicted to show an activity cliff with a potent molecule already in the dataset.
\end{abstract}

@@

The landscape paradigm for SAR data was first proposed 20 years ago \cite{Johnson:1990ys} and has
recently seen a resurgence with a number of studies describing new ways to quantify and visualize
activity landscapes. When SAR data is viewed as a landscape, with the X-Y plane representing
structural characteristics (which will usually be a 2-dimensional representation of a
multi-dimensional descriptor space) and the Z-axis representing the observed activities, one can
identify two broad types of regions on the landscape - smooth rolling regions, corresponding to set
of molecules exhibiting continuous SAR (i.e., similar structures and similar activities) and rough,
gorge-like regions (i.e., very similar structures, but large differences in activity) corresponding
to molecules that exhibit SAR discontinuity. The latter have also been term activity cliffs
\cite{Maggiora:2006aa}. From a medicinal chemistry point of view, the latter regions of a landscape
can be the most interesting as they can provide insight into structural features that are key to
improving (or conversely reducing) potency. There is a rich history of methods that have correlated
structural differences with corresponding differences in activity -- matched molecular pairs
\cite{Leach:2006aa}, SAS maps \cite{Maggiora:2011kx} and more recently SALI \cite{Guha:2008aa} and
SARI \cite{Peltason:2007aa}. Both SALI and SARI focus on numerically characterizing a structure
activity landscape. The former is defined for a pair of molecules as
\begin{equation}
  \label{eq:1}
  S_{i,j} = \frac{|A_i - A_j|}{1 - sim(i,j)}
\end{equation}
where $A_i$ and $A_j$ represent the observed activities of molecules $i$ and $j$, and $sim(i,j)$
represents the structural similarity between the two molecules (usually based on some form of
fingerprint similarity). Using the SALI metric, one can take a collection of $n$ molecules and
represent them as an $n \times n$ matrix of SALI values - larger values representing more
significant activity cliffs. The SARI approach is based on a score defined as
\begin{equation}
  \label{eq:2}
  \mathrm{SARI} = \frac{1}{2} \left(\mathrm{score}_{cont} + (1 -
    \mathrm{score}_{disc}) \right)
\end{equation}
where the individual score terms are derived on the basis of potency and pairwise similarities. The
reader is referred to Ref. \citen{Peltason:2007aa} for a detailed discussion of this approach.

@@

\section{Datasets}            
\label{sec:datasets}
For this study we considered a number of datasets, which are summarized in Table \ref{tab:data}. The
Cavalli dataset was employed in our previous studies consisted of 30 molecules studied by Cavalli et
al\cite{Cavalli:2002aa} as possible hErg inhibitors using a pharmacophore modeling approach. The
remaining datasets were obtained from ChEMBL. All three assays involved direct binding to a human
target and we considered the subset of molecules in each assay that had non-censored experimental
values. The Costanzo dataset\cite{Costanzo:2005uq} consisted of 60 $\alpha$-ketoheterocyclic
inhibitors of $\alpha$-thrombin. The reported $\mathrm{IC_{50}}$ values ranged from 3 nM to 82
$\mu$M. The Kalla dataset\cite{Kalla:2006kx} consisted of 38 8-(C-4-pyrazolyl) xanthines, identified
as antagonists of the A2B adenosine receptor. The reported $K_i$ values ranged from 0.9 nM to 42
$\mu$M. Finally, the Dai dataset\cite{Dai:2007kl} consisted of 44 3-aminoindazole derivatives
studied for their ability to inhibit the VEGF and PDGF receptor families, with $\mathrm{IC_{50}}$
values ranging from 3 nM to 12~$\mu$M.

@@

\section{Methodology}
\label{sec:methodology}
As noted above, the problem of identifying activity cliffs involving new molecules can be reduced to
predicting the SALI value for the new molecule and a pre-existing molecule. Note that this does not
result in an activity prediction for the new molecule; rather, it allows us to rank a set of new
molecules in terms of their predicted ability to exhibit a significant activity difference from one
or more of the molecules in the training set.

Given a training set of $N$ molecules, we generate a new training set
of $\frac{N(N-1)}{2}$ objects, where each object is a pair of
molecules from the original training set. The dependent variable for
each pair $i,j$, is the SALI value, $S_{i,j}$. SALI values were
evaluated using the 1051-bit BCI keyed fingerprints or the CDK
1024-bit path fingerprints. Based on the
definition of SALI, it is possible that a pair of molecules have a
$T_c = 1.0$ resulting in infinite values. For such cases, we replaced
the infinite value with the highest non-infinite SALI value for that
dataset.

@@

Given the independent and dependent variables for the pairwise
dataset, we can now proceed to model development. For this study we
focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models can be proven to not
overfit and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature
selection step and work directly with the descriptor pool (after
removal of correlated and constant descriptors). Furthermore, there is
no reason, \emph{a priori} to assume that the underlying SARs are
linear. A random forest model, being an algorithmic
approach\cite{Breiman:2001nx} (as opposed to a distributional one such
as linear regression) makes no such assumptions. We employed the  
implementation of random forest from the \texttt{randomForest}
package, in R 2.11.0\cite{r}, using the defult settings.

In this work, we employed the CDK to evaluate 109 2D and
constitutional descriptors for each individual molecule. For each
dataset, we performed objective feature selection by removing
descriptors with constant or near-constant values followed by removal
of descriptors that are highly correlated with others (using an $R^2$
cutoff of 0.8). The size of the final descriptor pools is summarized
in Table \ref{tab:data}.

One feature of this approach is that one can build relatively robust
models even for datasets of small size - say, 20 molecules. Of course,
a large training set is just one aspect of a reliable model and other
considerations such as diversity, descriptor selection and so on still
play an important role.

@@

One question that arises from the preceding discussion is the
possibility that we are duplicating information between the dependent
and independent variables. More specifically, both the dependent
variable (i.e., the SALI values) and the aggregated descriptors
characterize the molecules' structure. One might therefore ask whether
models built using such data perform over-optimistically. Given the
nature of the descriptors used in the indepdent and dependent
variables (the former being based on multiple atom and bond features
and the latter derived from purel topological paths), we believe that such a
phenomenon is unlikely.  Fig.~\ref{fig:desccor1} displays a histogram
of the pairwise Pearson correlations between the SALI values and each
of the descriptor values for the Cavalli\cite{Cavalli:2002aa} dataset,
using the different aggregation functions described above. It is
evident, that the highest $R^2$, between any of the descriptors and
the SALI values is less than 0.15. Similar behavior was observed for
all the datasets used in this study. These observations
suggest that the problem of including correlated structural
information simultaneously in the dependent and independent variables
is minimal.

\section{Results}
\label{sec:applications}
We first consider the application of the activity cliff prediction
methodology to the Cavalli dataset. We developed three random forest
models, using the three aggregation descriptor aggregation functions
described in Section
\ref{sec:methodology}. Fig.~\ref{fig:rfmodel-cavalli} presents the
predicted versus observed SALI values from the three models and Table
\ref{tab:rfmodel-results} summarizes the performance metrics for the
three models. Overall the different aggregation functions do not
differ dramatically in terms of the final model performance, though
$f_{\textrm{mean}}$ appears to lead to the best performance.

While the models are certainly not the most accurate models, the
RMSE's and $R^2$ values are reasonable. Note that the predictions on
the low end of the SALI values are not as important as those at the
high end - simply because low SALI values correspond to small activity
cliffs, which are likely not very interesting. Given that observation,
it is encouraging to observe that all three models perform relatively
well at the higher end of the SALI spectrum. For this example, the
three most significant activity cliffs are in fact not very
significant activity cliffs in an absolute sense - the $T_c$ for the
the three pairs are just 0.2, 0.30 and 0.29, though the activity
differences were 4.93, 5.0 and 5.11 log units respectively.

We observed that the use of the geometric mean did not lead to models
that were any different from those generated using the arithmetic mean
as the aggregation function. This was also true of the other datasets
and so the following discussion ommits the results obtained when using
$f_{\textrm{geom}}$. Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and
$f_{\textrm{mean}}$ as the aggregation functions. Table
\ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the
aggregation function used, we see that for the Costanzo and Kalla
datasets, $f_{\textrm{diff}}$ leads to a slightly better model than
when using $f_{\textrm{mean}}$, though for the Dai dataset, the
difference is negligable.

Given the relatively small differences between models based on
different aggregation functions, we arbitrarily chose
$f_{\textrm{mean}}$ as the aggregagtion functin for subsequent
analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and
\ref{fig:dai-pred} summarize the performance of the models built using
the reported activity data and the pairwise SALI values for the
Costanzo, Kalla and Froimowitz datasets, respectively. As with the
Cavalli dataset, the SALI models exhibit more variance for small SALI
values (i.e., less significant cliffs) but appear to perform better at
the the higher end of the SALI values. However in all three cases, the
most significant cliffs exhibit significant variance. In comparison,
the models developed on the reported activity values are relatively
poor. This can be partially ascribed to the small sizes of the
dataset.

@@

Clearly, modeling the pairwise SALI values leads to reasonable models
the datasets discussed here. However, the primary goal of this
approach is to extend the the structure-activity landscape for a set
of molecules. In other words, can one use the model to identify
compounds that will exhibit an activity cliff with one or more
molecules in the dataset. Given a new molecule, one can evaluate the
descriptors for this molecule and then evaluate the pairwise
descriptor values with each member of the training set. Thus, given a
new molecule, we must obtain predictions for $n$ ``pairwise objects'',
where $n$ is the number of molecules in the training set. Given these
$n$ predictions one can proceed to identify training set molecules
with which the new molecule might exhibit a cliff.

We first performed this analysis with the Kalla dataset. Since
predictions of new molecules involve $n$ predictions per molecule, we
removed three molecules from the dataset and rebuilt the random
forest model. The hold out molecules, were specifically selected since
they displayed significant activity cliffs with various members of the
training set. We then evaluted the pairwise descriptor values for
these three with the training set molecules and obtained predictions
of the log(SALI) values. Fig.~\ref{fig:tset-pset-kalla}A
summarizes the performance of the model on the training set as well as
for the prediction set. From a numerical point of view, the
performance of the model on he prediction set degrades somewhat (RMSE
of 0.51 versus 0.41 for the training set). As noted above, the
variance of the prediction is higher at the lower scale of log(SALI)
values. However, at the higher end of the scale there a number of
pairs that are well predicted. Yet, it is clear that a number of
relatively significant activity cliffs have been underestimated.

Given that large SALI values can arise due to high structural
similarity, even when the activity difference is small, simply
examining log(SALI) values is not comletely informative. Specifically,
we are interested in the predictions for the actual activity cliffs,
where both the difference in activity and the structural similarity is
very high. Fig.~\ref{fig:tset-pset-kalla}B visualizes this
information. Each point coresponds to a pair of molecules, which are
shaded by the absolute prediction residual for that pair. Thus, the
``true'' cliffs are represented by points lying towards the bottom
right corner of the plot. In this case we see that there are a number
of pairs of molecules with $1 - T_C \leq 0.1$ and 100-fold or better
difference in activity. While many of these pairs are associated with
a high prediction residual, there are a number with low to medium
residuals.

Fig.~\ref{fig:kalla-comparison} displays the structures of the three
hold out molecules (top row) and a selection of structures, which
which these hold out molecules were predicted to exhibit an activity
cliff. For each molecule, the $K_i$ value is listed along with the
absoute residual value (log(SALI) units). For molecules
\textbf{349288} and \textbf{349699}, the predicted cliffs are
relatively accurate, as evidenced by the low residuals. For
\textbf{349138} on the other hand, most of the residuals were were
relatively high. While we have highlighted a number of cliffs,
Fig.~\ref{fig:tset-pset-kalla}C makes it clear that the predictions
for this molecule were, on average, worse than for the other two
molecules.

Fig.~\ref{fig:tset-pset-dai} summarizes the results for the same set
of analyzes described above, applied to the Dai dataset. As before we
removed three molecules (top row of Fig.~\ref{fig:dai-comparison}) and
rebuilt the random forest on the paired cases with the remaining 41
molecules. We then predicted the pairwise SALI values for the three
hold-out molecules with the remaining 41 molecules (i.e., the
prediction set). A plot of the predicted versus observed log(SALI)
values for the training and prediction sets are displayed in
Fig.~\ref{fig:tset-pset-dai}A. As before the model performs poorly on
the lesser cliffs, but improves for the more significant cliffs (RMSE
for the training set and prediction sets were 0.37 and 0.43 log units
respectively). However, the 2 most significant cliffs are
under-estimated. However, in comparison to the Kalla dataset, the
distribution of absolute residuals aggregated by the hold-out molecule
(Fig.~\ref{fig:tset-pset-dai}C) are reatively similar, with a median
absolute residual of less than 0.3 log units.  For this dataset, the
number of ``real'' cliffs is relatively low as shown in
Fig.~\ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that
display moderate to significant cliffs are relatively well predicted -
in fact, the maximum residuals are observed for the least significant
cliffs.

>>>>

\maketitle
\begin{abstract}
  The notion of activity cliffs is an intuitive approach to characterizing structural features that
  play a key role in modulating biological activity of a molecule. A variety of methods have been
  described to quantitatively characterize activity cliffs, such as SALI and SARI. However, these
  methods are primarily retrospective in nature; highlighting cliffs that are already present in the
  dataset. The current study focuses on employing a pairwise characterization of a dataset to train
  a model to predict whether a new molecule will exhibit an activity cliff with one or more members
  of the dataset. The approach is based on predicting a value for pairs of objects rather than the
  individual objects themselves (and thus allows for robust models even for small structure-activity
  relationship datasets). We extracted structure-activity data for several ChEMBL assays and
  developed random forest models to predict SALI values, from pairwise combinations of molecular
  descriptors. The models exhibited reasonable RMSE's though, surprisingly, performance on the more
  significant cliffs tended to be better than on the lesser ones. Our results indicate that the
  predictive models are able to prioritize molecules in terms of their ability to activity cliffs
  and can be used as a way to extend an observed activity landscape to identify molecules that could
  lead to significant improvements in activity. Conversely, the method can be used to deprioritize
  molecules that are predicted to show an activity cliff with a potent molecule already in the
  dataset.
\end{abstract}

@@

The landscape paradigm for structure-activity relationship (SAR) data was first proposed 20 years
ago \cite{Johnson:1990ys} and has recently seen a resurgence with a number of studies describing new
ways to quantify and visualize activity landscapes. When SAR data is viewed as a landscape, with the
X-Y plane representing structural characteristics (which will usually be a 2-dimensional
representation of a multi-dimensional descriptor space) and the Z-axis representing the observed
activities, one can identify two broad types of regions on the landscape - smooth rolling regions,
corresponding to set of molecules exhibiting continuous SAR (i.e., similar structures and similar
activities) and rough, gorge-like regions (i.e., very similar structures, but large differences in
activity) corresponding to molecules that exhibit SAR discontinuity. The latter have also been term
activity cliffs \cite{Maggiora:2006aa}. From a medicinal chemistry point of view, the latter regions
of a landscape can be the most interesting as they can provide insight into structural features that
are key to improving (or conversely reducing) potency. There is a rich history of methods that have
correlated structural differences with corresponding differences in activity -- matched molecular
pairs \cite{Leach:2006aa}, SAS maps \cite{Maggiora:2011kx} and more recently SALI \cite{Guha:2008aa}
and SARI \cite{Peltason:2007aa}. Both SALI and SARI focus on numerically characterizing a structure
activity landscape. The former is defined for a pair of molecules as
\begin{equation}
  \label{eq:1}
  S_{i,j} = \frac{|A_i - A_j|}{1 - sim(i,j)}
\end{equation}
where $A_i$ and $A_j$ represent the observed activities of molecules $i$ and $j$, and $sim(i,j)$
represents the structural similarity between the two molecules (usually based on some form of
fingerprint similarity). Using the SALI metric, one can take a collection of $n$ molecules and
represent them as an $n \times n$ matrix of SALI values - larger values representing more
significant activity cliffs. The SARI approach is based on a score defined as
\begin{equation}
  \label{eq:2}
  \mathrm{SARI} = \frac{1}{2} \left(\mathrm{score}_{cont} + (1 -
    \mathrm{score}_{disc}) \right)
\end{equation}
where the individual score terms are derived on the basis of potency and pairwise similarities. The
reader is referred to Ref. \citen{Peltason:2007aa} for a detailed discussion of this approach.

@@

\section{Datasets}            
\label{sec:datasets}
For this study we considered a number of datasets, which are summarized in Table \ref{tab:data}. The
Cavalli dataset was employed in our previous studies consisted of 30 molecules studied by Cavalli et
al\cite{Cavalli:2002aa} as possible hErg inhibitors using a CoMFA modeling approach. The reported
endpoint for the molecules was a pIC$_{50}$.The remaining datasets were obtained from ChEMBL. All
three assays involved direct binding to a human target and we considered the subset of molecules in
each assay that had non-censored experimental values. The Costanzo dataset\cite{Costanzo:2005uq}
consisted of 60 $\alpha$-ketoheterocyclic inhibitors of $\alpha$-thrombin. The reported
$\mathrm{IC_{50}}$ values ranged from 3 nM to 82 $\mu$M. The Kalla dataset\cite{Kalla:2006kx}
consisted of 38 8-(C-4-pyrazolyl) xanthines, identified as antagonists of the A2B adenosine
receptor. The reported $K_i$ values ranged from 0.9 nM to 42 $\mu$M. Finally, the Dai
dataset\cite{Dai:2007kl} consisted of 44 3-aminoindazole derivatives studied for their ability to
inhibit the VEGF and PDGF receptor families, with $\mathrm{IC_{50}}$ values ranging from 3 nM to
12~$\mu$M.

@@

\section{Methodology}
\label{sec:methodology}
As noted above, the problem of identifying activity cliffs involving new molecules can be reduced to
predicting the SALI value for the new molecule and a preexisting molecule. Note that this does not
result in an activity prediction for the new molecule; rather, it allows us to rank a set of new
molecules in terms of their predicted ability to exhibit a significant activity difference with
respect to one or more of the molecules in the training set.

Given a training set of $N$ molecules, we generate a new training set of $\frac{N(N-1)}{2}$ objects,
where each object is a pair of molecules from the original training set. The dependent variable for
each pair $i,j$, is the SALI value, $S_{i,j}$. SALI values were evaluated using the 1051-bit BCI
keyed fingerprints (Digital Chemistry, UK) or the CDK\cite{Steinbeck:2006aa,Steinbeck:2003bh}
1024-bit path fingerprints. Based on the definition of SALI, it is possible that a pair of molecules
have a $T_c = 1.0$ resulting in infinite values. For such cases, we replaced the infinite value with
the highest non-infinite SALI value for that dataset.

@@

Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models tend not to overfit the training data (given a
sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature selection step and work directly
with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
linear regression) makes no such assumptions. We employed the implementation of random forest from
the \texttt{randomForest} package, in R 2.11.0\cite{r}, using the default settings.

In this work, we employed the CDK to evaluate 109 2D and constitutional descriptors for each
individual molecule. For each dataset, we performed objective feature selection by removing
descriptors with constant or near-constant values followed by removal of descriptors that are highly
correlated with others (using an $R^2$ cutoff of 0.8). The sizes of the final descriptor pool for
each dataset are summarized in Table \ref{tab:data}.

A side effect of the proposed approach is that one can build relatively robust models for datasets of
small size -- say, 20 molecules. Of course, a large training set is just one aspect of a reliable
model and other considerations such as diversity, descriptor selection and so on still play an
important role.

@@

The preceding discussion raises the issue of explicitly duplicating structural information in the
independent and dependent variables. More specifically, both the dependent variable (i.e., the SALI
values) and the aggregated descriptors characterize the molecules' structure. One might therefore
ask whether models built using such data perform over-optimistically. Given the nature of the
descriptors used in the independent and dependent variables (the former being based on multiple atom
and bond features and the latter derived from purely topological paths), we believe that such a
phenomenon is unlikely.  Fig.~\ref{fig:desccor1} displays a histogram of the pairwise Pearson
correlations between the SALI values and each of the descriptor values for the
Cavalli\cite{Cavalli:2002aa} dataset, using the different aggregation functions described above. It
is evident, that the highest $R^2$, between any of the descriptors and the SALI values is less than
0.15. Similar behavior was observed for all the datasets used in this study. These observations
suggest that problems arising from the inclusion of structural information simultaneously in the
dependent and independent variables is minimal.

\section{Results}
\label{sec:applications}
We first consider the application of the SALI prediction methodology to the Cavalli dataset. We
developed three random forest models, using the three aggregation descriptor aggregation functions
described in Section \ref{sec:methodology}. Fig.~\ref{fig:rfmodel-cavalli} presents the predicted
versus observed SALI values from the three models and Table \ref{tab:rfmodel-results} summarizes the
performance metrics for the three models. Overall the different aggregation functions do not differ
dramatically in terms of the final model performance as root mean square error (RMSE) ranges from
0.98 to 1.11 log units. However, $f_{\textrm{mean}}$ appears to lead to the best performance (RMSE = 0.98 and
$R^2$ = 0.67). In contrast, the RMSE and $R^2$ values for Y-scrambled\cite{Rucker:2007aa} models
using the different aggregation functions ranged from 1.58 to 1.61 log units and 0.003 to 0.01, respectively.

Note that the predictions on the low end of the SALI values are not as important as those at the
high end - simply because low SALI values correspond to small activity cliffs, which are likely not
very interesting. Given that observation, it is encouraging to observe that all three models perform
relatively well at the higher end of the SALI spectrum. For this example, the three most significant
activity cliffs are in fact not very significant activity cliffs in an absolute sense - the $T_c$
for the three pairs are just 0.2, 0.30 and 0.29, though the activity differences were 4.93, 5.0 and
5.11 log units respectively.

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. Fig.~\ref{fig:rfmodel-chembl} summarizes the models built on the ChEMBL
datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the aggregation functions. Table
\ref{tab:rfmodel-results} reports the model statistics.  While the differences in model performance
do not vary significantly with the aggregation function used, we see that for the Costanzo and Kalla
datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms of RMSE and $R^2$ than when
using $f_{\textrm{mean}}$.

Given the relatively small differences between models based on different aggregation functions, we
arbitrarily chose $f_{\textrm{mean}}$ as the aggregation function for subsequent
analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred} summarize the
performance of the models built using the reported activity data and the pairwise SALI values for
the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI models
exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
better at the higher end of the SALI values. However in all three cases, the most significant cliffs
exhibit significant variance. In comparison, the models developed on the reported activity values
are relatively poor. This can be partially ascribed to the small sizes of the dataset. It should be
note that this methodology is not aimed at modeling the original, usually small, SAR dataset. It is
well known (and evident from the current analyses) that small datasets do not lead to very robust or
reliable machine learning models. Instead, the methodology focuses on the \emph{pairwise} dataset --
which are usually large enough for a statistically significant model.

@@

While predictive modeling of the SALI values is a first step, the primary goal of this approach is
to extend the structure-activity landscape for a set of molecules. In other words, can one use
the model to identify compounds that will exhibit an activity cliff with one or more molecules in
the dataset? Given a new molecule, one can evaluate the descriptors for this molecule and then
evaluate the pairwise descriptor values with each member of the training set. Thus, given a new
molecule, we must obtain predictions for $n$ ``pairwise objects'', where $n$ is the number of
molecules in the training set. Given these $n$ predictions one can proceed to identify training set
molecules with which the new molecule might exhibit a cliff.

We first performed this analysis with the Kalla dataset. Since predictions of new molecules involve
$n$ predictions per molecule, we removed three molecules from the dataset and rebuilt the random
forest model. The hold out molecules, were specifically selected since they displayed significant
activity cliffs with various members of the training set. We then evaluated the pairwise descriptor
values for these three with the training set molecules and obtained predictions of the log(SALI)
values. Fig.~\ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training set
as well as for the prediction set. From a numerical point of view, the performance of the model on
he prediction set degrades somewhat (RMSE of 0.51 versus 0.41 for the training set). As noted above,
the variance of the prediction is higher at the lower scale of log(SALI) values. However, at the
higher end of the scale there a number of pairs that are well predicted. Yet, it is clear that a
number of relatively significant activity cliffs have been underestimated.

Given that large SALI values can arise due to high structural similarity, even when the activity
difference is small, simply examining log(SALI) values is not completely informative. Specifically,
we are interested in the predictions for the actual activity cliffs, where both the difference in
activity and the structural similarity is very high. Fig.~\ref{fig:tset-pset-kalla}B visualizes this
information. Each point corresponds to a pair of molecules, which are shaded by the absolute
prediction residual for that pair. Thus, the ``true'' cliffs are represented by points lying towards
the bottom right corner of the plot. In this case we see that there are a number of pairs of
molecules with $1 - T_C \leq 0.1$ and 100-fold or better difference in activity. While many of these
pairs are associated with a high prediction residual, there are a number with low to medium
residuals.

Fig.~\ref{fig:kalla-comparison} displays the structures of the three hold out molecules (top row)
and a selection of structures, which which these hold out molecules were predicted to exhibit an
activity cliff. For each molecule, the $K_i$ value is listed along with the absolute residual value
(log(SALI) units). For molecules \textbf{349288} and \textbf{349699}, the predicted cliffs are
relatively accurate, as evidenced by the low residuals. For \textbf{349138} on the other hand, most
of the residuals were were relatively high. While we have highlighted a number of cliffs,
Fig.~\ref{fig:tset-pset-kalla}C makes it clear that the predictions for this molecule were, on
average, worse than for the other two molecules.

Fig.~\ref{fig:tset-pset-dai} summarizes the results for the same set of analyzes described above,
applied to the Dai dataset. As before we removed three molecules (top row of
Fig.~\ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining
41 molecules. We then predicted the pairwise SALI values for the three hold-out molecules with the
remaining 41 molecules (i.e., the prediction set). A plot of the predicted versus observed log(SALI)
values for the training and prediction sets are displayed in Fig.~\ref{fig:tset-pset-dai}A. As
before the model performs poorly on the lesser cliffs, but improves for the more significant cliffs
(RMSE for the training set and prediction sets were 0.37 and 0.43 log units respectively). However,
the 2 most significant cliffs are under-estimated. However, in comparison to the Kalla dataset, the
distribution of absolute residuals aggregated by the hold-out molecule
(Fig.~\ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than
0.3 log units.  For this dataset, the number of ``real'' cliffs is relatively low as shown in
Fig.~\ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to
significant cliffs are relatively well predicted - in fact, the maximum residuals are observed for
the least significant cliffs.

We then considered the three hold-out molecules and some of the members of the training set with
which they are predicted to show activity cliffs. For this dataset, there are relatively few
``true'' activity cliffs. For example, while molecules \textbf{371259} and \textbf{371307} have a
1000-fold difference in activity, their $T_c = 0.68$, and would not represent a truly significant
cliff. However, the pair is well predicted with a residual of 0.04 log units. if we consider the
subset of predictions where the $T_c \geq 0.8$ we note that the median absolute residual is 0.24 log
units. However, the activity differences are not always significant ranging from 1.1-fold to
13-fold. Two such examples are shown in Fig.~\ref{fig:dai-comparison}, where the only difference is
in the position of the methyl substituent. Similar behavior is observed with the other hold-out
molecules.
